A Systematic Literature Review on Extensions to Role-Based Access Control

### 2013-05-01 paper/sections/discussion.tex
address comments
John's comments

\subsection{Metrics for the Evaluation of Extensions to RBAC}

@@

\begin{itemize}
	\item \textbf{Ease of Privilege Assignments}: measures number of steps for assigning, changing, or deleting a privilege for users and roles when a user or administrator manages RBAC. If a model requires more steps for such a given task, there is a potential increase in human or system error due to increased complexity. While extended RBAC may have similar features to that of the RBAC standard model, we measure worst cases after extended, and the gap between an extension model and the RBAC standard model.

@@

While the concept of the RBAC model is clear and can be applicable to any system, it is challenging to implement the security mechanism of RBAC because the RBAC model is can be interpreted in more than one interpretation due to its complexity. In order to bridge gap between RBAC standard and its implementation, NIST published RBAC Implementation and Interoperability standard (RIIS/ ANSI INCITS-459) in 2011. This standard specifies how to implement RBAC system, which is consistent with NIST RBAC standard. Moreover, this standard describes interoperability specification where one RBAC implementation can be translated to another one.

>>>>

\subsection{Metrics for the Evaluation of RBAC Extensions}

@@

\begin{itemize}
	\item \textbf{Ease of Privilege Assignments}: measures number of steps for assigning, changing, or deleting a privilege for users and roles when a user or administrator manages RBAC. If a model requires more steps for such a given task, there is a potential increase in human or system error due to increased complexity. While extended RBAC may have similar features to that of the RBAC$_{m}$, we measure worst cases after extended, and the gap between an extension model and the RBAC$_{m}$.

@@

While the concept of the RBAC$_{m}$ is clear and can be applicable to any system, it is challenging to implement the security mechanism of RBAC because the RBAC$_{m}$ is can be interpreted in more than one interpretation due to its complexity. In order to bridge gap between RBAC standard and its implementation, NIST published RBAC Implementation and Interoperability standard (RIIS/ ANSI INCITS-459) in 2011. This standard specifies how to implement RBAC system, which is consistent with NIST RBAC standard. Moreover, this standard describes interoperability specification where one RBAC implementation can be translated to another one.

### 2013-05-01 paper/sections/intro.tex
fixing after spell check and grammar check based on word
spell check and grammar check based on word
address comments
John's comments

Role based access control (RBAC) is used for maintaining and managing an organization's 
access control based on assigning permissions to roles, and roles to users, instead of 
assigning the permissions directly to individual users. 
The use of an RBAC model has become popular since the National Institute for Standards and Technology (NIST) 
first proposed the RBAC standard model in the 1990s \cite{o20102010}. The RBAC standard model is widely used for securing 
various applications including web services, database applications, and healthcare applications. 
In the RBAC standard model, roles represent a set of permissions needed to perform a particular job function within an organization.  
One to many users who are involved in that specific job function within the organization can then be 
assigned to the role to inherit the required access instead of assigning individual permissions to individual users. 
The ability to logically group both users and permissions into smaller total entities becomes paramount for managing access control
as an organization grows and the number of permissions and users scales upward, 
The RBAC standard model has advantages for managing access control. For example, if a user requires access to resources 
associated with the manager role within a given organization, security policy administrators need only associate the user with the manager role.

The RBAC standard was first proposed in the 1990s when NIST 
requested that a unified standard be created by combining the Ferraiolo and Kuhn model \cite{ferraiolokuhn} with the framework 
proposed by Sandhu, et al \cite{sandhu1996role}.  
In 2004, this standard was adopted as ANSI/INCITS 359-2004 approved by American National Standards Institute (ANSI) and the InterNational Committee for Information Technology Standards (INCITS). 
The development of a standard was inspired by an economic impact study done during the 1990s and later confirmed in 2010 \footnote{\url{http://csrc.nist.gov/groups/SNS/rbac/documents/20101219_RBAC2_Final_Report.pdf}} that showed the cost savings of RBAC implementation and maintenance. 
Prior to the development of the RBAC standard, vendors proposed and implemented their own defined RBAC features without general agreement on a unified definition of RBAC or feature set. 
The RBAC standard addressed this issue by providing definitions and a functional specification of RBAC features using a reference model.

Innovation and the use of RBAC in new domains has led to scenarios that, by some accounts, the standard RBAC model cannot handle \cite{kuhn2010adding}.  
For example, Ni, Q. et al \cite{ni2010privacy} state that RBAC is ``not designed to enforce privacy policies and barely 
meet privacy protection requirements'' with the introduction of privacy concerns in domains such as healthcare. 
Since the introduction of the NIST RBAC standard in 2000, extension models to the RBAC standard have appeared in the literature, each adding one or more features on top of the core RBAC model. 
A cursory examination of the extension model papers does not provide good comparisons for developers and researchers to build upon. 
Further, these extension models are each building upon and adding features to a standard that was designed to reduce the economic impact experienced by enterprises and increase interoperability \cite{o20102010}.

\textit{The goal of our work is to aid practitioners and researchers in choosing an extension to the RBAC standard model, and in understanding
how extensions to RBAC are evaluated by providing an assessment of the state of extensions to the RBAC standard model. We accomplished this through: 
establishment of a set of extension categories, an examination of the state of the art in evaluations of extension models and a breakdown of the 
motivations that have led to extension of the RBAC standard model.} With respect to our goal, the authors addressed the following research questions:

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item RQ1: What categories exist within extensions to RBAC?
\item RQ2: What are the motivations behind extensions to RBAC?
\item RQ3: Do the extensions to RBAC have corresponding implementations?
\item RQ4: How are extensions to RBAC evaluated theoretically and in practice?
\item RQ5: What domains have extensions to RBAC been created for?
\item RQ6: What commonalities or generalizations exist across all categories?
\end{itemize}

The authors performed a systematic literature review designed to explore the current body of research in the area of extensions to the RBAC standard model. 
The review we performed yielded 1716 papers, of which 28 were deemed primary sources for inclusion as extension models to the RBAC standard model.  
This review is intended to serve as a starting place for researcher's looking to tackle new problems in the realm of authorization and a reference guide for discovering
the current state of extensions to the RBAC standard model. 
For developer's looking to find a model to fit their potentially unique access control needs, this review will provide a basis for comparison and easy look-up for what extension model to use.  
Further, the review provides insight into the state of the art in evaluating RBAC extension models.

Our research provides contributions to the community in the following ways:

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item Summarizes current research on extensions to RBAC.
\item Guides a direction for a new standard based on an extension to the RBAC standard model.
\end{itemize}

The rest of the paper is organized as follows. 
Section 2 covers background and the core RBAC model. 
Section 3 lays out the process used in conducting the review. 
Sections 4-9 present the analysis and discussion of the research questions. 
Section 10 is a general discussion section brings together all the research questions.
Section 11 contains the final conclusions.

>>>>

Role-Based Access Control (RBAC) is widely used for maintaining and managing an organization's 
access control based on assigning permissions to roles, and roles to users, instead of 
assigning the permissions directly to individual users.
RBAC is used for securing 
various applications including web services, database applications, and healthcare applications. 
In RBAC, roles represent a set of permissions needed to perform a particular job function within an organization.  
Multiple users who are involved in that specific job function within the organization can then be 
assigned to a single role to inherit the required access. 
The ability to logically group users into roles associated with permissions becomes paramount for managing access control as an organization grows
and the number of permissions and users scales upward.
As permissions can be managed by a role instead of a user, RBAC has an advantage to significantly reduce complexity of security
administration.
For example, if a user requires access to resources 
associated with a manager role within a given organization, security policy administrators need only associate the user with the manager role instead
of assigning permissions of the manager role to the user.

The use of RBAC has become popular since the National Institute for Standards and Technology (NIST) 
first proposed RBAC standard in the 1990s \cite{o20102010}.
The RBAC standard was first proposed when NIST 
requested that a unified standard be created by combining the Ferraiolo and Kuhn model \cite{ferraiolokuhn} with the framework 
proposed by Sandhu, et al \cite{sandhu1996role}.  
In 2004, this standard was adopted as ANSI/INCITS 359-2004 approved by American National Standards Institute (ANSI) and the InterNational Committee for Information Technology Standards (INCITS). 
The development of a standard was inspired by an economic impact study done during the 1990s and later confirmed in 2010 \footnote{\url{http://csrc.nist.gov/groups/SNS/rbac/documents/20101219_RBAC2_Final_Report.pdf}} that showed the cost savings of RBAC implementation and maintenance. 
Prior to the development of the RBAC standard, vendors proposed and implemented their own defined RBAC features without general agreement on a unified definition of RBAC or feature set. The RBAC standard includes RBAC reference models (RBAC$_{m}$ in short). RBAC$_{m}$ serve as a basis for defining scope of RBAC features and functional specifications of RBAC features.

Innovation and the use of RBAC in new domains has led to scenarios that, by some accounts, the RBAC$_{m}$ cannot handle \cite{kuhn2010adding}.  
For example, Ni et al. \cite{ni2010privacy} state that RBAC$_{m}$ are ``not designed to enforce privacy policies and barely 
meet privacy protection requirements'' with the introduction of privacy concerns in domains such as healthcare. 
Since the introduction of the RBAC standard, extension models to the RBAC$_{m}$ have appeared in the literature, each adding one or more features on top of components in the RBAC$_{m}$.
%A cursory examination of the extension model papers does not provide good comparisons for developers and researchers to build upon. 
Further, these extension models are each building upon and adding features to a standard that was designed to reduce the economic impact experienced by enterprises and increase interoperability \cite{o20102010}.

\textit{The goal of our work is to aid practitioners and researchers in choosing an extension to the RBAC$_{m}$, and in understanding
how extensions to RBAC$_{m}$ are evaluated by providing an assessment of RBAC extensions. We accomplished this through: 
establishment of a set of RBAC extension categories, an examination of the state of the art in evaluations of RBAC extension, and a breakdown of the 
motivations that have led to extensions to RBAC$_{m}$.} To accomplish this goal, we seek to answer following research questions:

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item RQ1: What categories exist within extensions to RBAC$_{m}$?
\item RQ2: What are the motivations behind extensions to RBAC$_{m}$?
\item RQ3: Do the extensions to RBAC$_{m}$ have corresponding implementations?
\item RQ4: How are extensions to RBAC$_{m}$ evaluated theoretically and in practice?
\item RQ5: What domains have extensions to RBAC$_{m}$ been created for?
\item RQ6: What commonalities or generalizations exist across all categories?
\end{itemize}

We performed a systematic literature review designed to explore the current body of research in the area of extensions to the RBAC$_{m}$. 
The review we performed yielded 1,716 papers, of which 28 were deemed primary sources for inclusion as extension models to the RBAC$_{m}$.  
This review is intended to serve as a starting place for researchers looking to tackle new problems in the realm of authorization and as a reference guide for discovering
the current state of extensions to the RBAC$_{m}$. 
For developers looking to find a model to fit their potentially unique access control needs, this review provides a basis for comparison and easy look up for what extension model to use.  
Furthermore, the review provides insight into the state of the art in evaluating RBAC extension models.

Our research provides contributions to the community in following ways:

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item Summarizes current research on extensions to RBAC$_{m}$.
\item Guides a direction for a new standard based on an extension to the RBAC$_{m}$.
\end{itemize}

The rest of the paper is organized as follows. 
Section~\ref{sec:core-rbac} presents background and the RBAC standard. 
Section~\ref{sec:process} presents methodology and process, which we used in conducting the systemic literature review. 
Sections~\ref{sec:categorization}-\ref{sec:generalizations} present analysis and discussion of the research questions. 
Section~\ref{sec:discussion} discusses issues about extensions to RBAC$_{m}$.
Section~\ref{sec:conclusion} concludes the paper.

### 2013-05-01 paper/sections/process.tex
address comments
John's comments

The authors adopted and applied a systematic literature review following recommendations from Kitchenham's suggested processes \cite{kitchenham2007guidelines}.  The systematic literature review process was broken down in to four stages and the rest of this section is broken down by each stage.  The stages were:

@@

For the first phase of our systematic literature review, the authors developed a search strategy for finding papers.  The search strategy was executed by an automated comprehensive search taking as input a set of academic search engines and a list of search terms.  The search was performed by applying each search term to each engine incrementally until stopping criteria were met. Table \ref{tab:search_results} lists the four search engines along the left most column with the three search term across the top row along with the papers for each criteria and engine combination.  The search algorithm was performed as follows:

@@

The result set size stopping criteria was chosen due to a technical limitation of some search engines.  The stopping criteria related to the last ten titles is meant to stop after relevant results are no longer being returned by the search engine.  After gathering all 12 data sets, the authors combined the papers into a master list by systematically comparing the bibliographic information for each.  The master list of 1716 papers was used to perform a series of elimination rounds to narrow the list of papers and identify primary sources.  Table \ref{tab:eliminations} shows the total number of papers selected by each author for each round and how many papers from the disjoint set for each round survived to the next round.

@@

The elimination rounds were performed based on reading of the title, abstract and finally the paper itself.  While each elimination stage had a unique set of criteria for elimination the general procedure for elimination for the researchers was as follows.

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item The two first authors independently classified papers as relevant, irrelevant or uncertain based off elimination criteria
\item Those papers marked as relevant by both reviewers were kept and those marked irrelevant by both were thrown out.
\item Papers marked as relevant, or irrelevant by a single reviewer were combined with all papers marked as uncertain and discussed by both reviewers.  From this discussion, papers were either thrown out or kept until the next round of the review.
\end{itemize}

@@

The first round of elimination was performed by strict examination based on the title.  Each author was tasked with deciding on elimination by answering the following questions:

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item Did the title contain a reference to 'role based access control' or 'RBAC'? 
\item Did the title contain a reference to 'model'?
\end{itemize}

@@

The second round of elimination was based on strict reading of the abstracts of papers that survived title elimination.  Researchers read each abstract and evaluated relevancy based off:

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item Does the abstract mention a proposed model?
\item Does the abstract mention extension of role-based access control?
\item Does the abstract mention an implementation, evaluation or domain for their model?
\end{itemize}

@@

The final elimination round involved taking the entire paper into consideration and answering five questions that would serve as the basis for elimination.  The data collected by answering these questions served as the basis towards answering the research questions.  The questions each reviewer attempted to answer based on the content of the paper was:

\begin{enumerate}
\setlength{\itemsep}{0.25pt}
\item Does this model extend the core model? (Exclusion)
\item Do the researchers give evidence that RBAC needs extension? (Inclusion)
\item Was the paper and subsequent model inspired by a real world example?  (Conditional Inclusion)
\item Did the researchers offer any evaluation of the proposed model? If yes, how did they do one? If no, why? (Conditional Inclusion)
\item Did the authors implement their model? (Inclusion)
\end{enumerate}

Question 1 was a definitive exclusion criterion as any paper that failed in the affirmative was rejected.  Questions 3 and 4 were marked as conditional includes given that they were connected in making a decision.  A paper that met Question 3 but not 4, or met 4 but not 3 would be included since in some cases the real world examples served as research evaluations and without this conditional include the paper list size would be too small to be significant.

@@

\multicolumn{1}{ |c| }{Eric}  &  & 305 & 86 & 46 \\ \hline
\multicolumn{1}{ |c| }{Hwang} &  & 176 & 102 & 42 \\ \hline

@@

@@

After selection of primary sources, the next step was to extract data from each paper that pertained to our research questions in order to look for trends.  The first step was to take the individual data generated from the final elimination round and organize this information around the research questions.  During the paper reading round, and resulting data, the fact that the papers were logically falling into a number of categorizations became evident.  Thus, the first step undertaken was to answer the question of what categories exist for the RBAC model extensions and what papers feel into what categories.

>>>>

We adopted and applied a systematic literature review following recommendations from Kitchenham's suggested processes \cite{kitchenham2007guidelines}.  The systematic literature review process was broken down into four stages and the rest of this section is broken down by each stage.  The stages were:

@@

For the first phase of our systematic literature review, we developed a search strategy for finding papers.  The search strategy was executed by an automated comprehensive search taking as input a set of academic search engines and a list of search terms.  The search was performed by applying each search term to each engine incrementally until the stopping criteria were met. Table \ref{tab:search_results} lists the four search engines along the left most column with the three search terms across the top row along with the papers for each criterion and engine combination.  The search algorithm was performed as follows:

@@

The result set size stopping criteria was chosen due to a technical limitation of some search engines.  The stopping criteria related to the last ten titles is meant to stop after relevant results are no longer being returned by the search engine.  After gathering all 12 data sets, we combined the papers into a master list, which includes only distinct papers by systematically comparing the bibliographic information for each.
%The master list of 1,716 papers was used to perform a series of elimination rounds to narrow the list of papers and identify primary sources.
Out of the master list of 1,716 papers, two reviewers, namely Reviewer 1 and Reviewer 2, conducted a series of elimination rounds to narrow the list of papers and identify primary sources. Table \ref{tab:eliminations} shows the total number of papers selected by each reviewer for each round and how many papers from the disjoint set for each round survived to the next round.

@@

The elimination rounds were conducted based on reading of the title, abstract, and finally the paper itself.  While each elimination stage had a unique set of criteria for elimination, the general procedure for elimination for the researchers was as follows.

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item The two first authors independently classified papers as relevant, irrelevant or uncertain based on elimination criteria
\item Those papers marked as relevant by both reviewers were kept and those marked irrelevant by both were thrown out.
\item Papers marked as relevant or irrelevant by a single reviewer were combined with all papers marked as uncertain and discussed by both reviewers.  From this discussion, papers were either thrown out or kept until the next round of the review.
\end{itemize}

@@

The first round of elimination was performed by examination based on the title.  Each author was tasked with deciding on elimination by answering the following questions:

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item Did the title contain a reference to 'role-based access control' or 'RBAC'? 
\item Did the title contain a reference to 'model'?
\end{itemize}

@@

The second round of elimination was based on reading of the abstracts of papers that survived title elimination.  Researchers read each abstract and evaluated relevancy based off:

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item Does the abstract mention a proposed model?
\item Does the abstract mention extension of role-based access control?
\item Does the abstract mention an implementation, evaluation, or domain for their model?
\end{itemize}

@@

The final elimination round involved reading the entire paper into consideration and answering five questions that would serve as the basis for elimination.  The data collected by answering these questions served as the basis towards answering the research questions.  Each reviewer seeks to answer following questions based on the content of the paper:

\begin{enumerate}
\setlength{\itemsep}{0.25pt}
\item Does this model extend the RBAC$_{m}$? (Exclusion)
\item Do the researchers give evidence that RBAC$_{m}$ needs extension? (Inclusion)
\item Was the paper and subsequent model inspired by a real world example?  (Conditional Inclusion)
\item Did the researchers offer any evaluation of the proposed model? If yes, how did they do one? If no, why? (Conditional Inclusion)
\item Did the authors implement their model? (Inclusion)
\end{enumerate}

Question 1 was a definitive exclusion criterion as any paper that failed in the affirmative was rejected.  Questions 3 and 4 were marked as conditional includes given that they were connected in making a decision.  A paper that met question 3 but not 4, or met 4 but not 3 would be included since in some cases the real world examples served as research evaluations and without this conditional include the paper list size would be too small to be significant.

@@

\multicolumn{1}{ |c| }{Reviewer 1}  &  & 305 & 86 & 46 \\ \hline
\multicolumn{1}{ |c| }{Reviewer 2} &  & 176 & 102 & 42 \\ \hline

@@

\end{tabular}
\label{tab:eliminations}
\end{table}

@@

After selection of primary sources, the next step was to extract data from each paper that pertained to our research questions in order to look for trends.  The first step was to take the individual data generated from the final elimination round and organize this information around the research questions.  During the paper reading round and resulting data, the fact that the papers were logically falling into a number of categorizations became evident.  Thus, the first step undertaken was to answer the question of what categories exist for extensions to RBAC$_{m}$ and what papers fell into what categories.
